---
title: "Filtering GTFS feeds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtering GTFS feeds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

GTFS feeds are often used to describe very large and complex public transport networks. These large files can become quite cumbersome to use, manipulate and move around, and it's not unfrequent that one wants to analyze a specific subset of the data.

**gtfstools** includes a few functions to filter GTFS feeds, thus allowing for faster and more convenient data processing. This vignette will introduce you to these functions and will cover their usage in detail.

We will start by loading the packages required by this demonstration into the current **R** session:

```{r, message = FALSE}
library(gtfstools)
library(ggplot2)
```

**gtfstools** includes the following filtering functions:

- `filter_by_trip_id()`
- `filter_by_route_id()`
- `filter_by_shape_id()`
- `filter_by_stop_id()`
- `filter_by_route_type()`
- `filter_by_service_id()`
- `filter_by_weekday()`
- `filter_by_time_of_day()`
- `filter_by_sf()`



## Filtering GTFS by `trip_id`, `route_id`, `shape_id`, `stop_id`, `route_type` or `service_id`:

The first five work in a very similar fashion. You specify a vector of identifiers, and the function keeps (or drops, as we'll see soon) all the entries that are in any way related to this id. Let's see how that works using `filter_by_trip_id()`:

```{r}
path <- system.file("extdata/spo_gtfs.zip", package = "gtfstools")
gtfs <- read_gtfs(path)
utils::object.size(gtfs)

head(gtfs$trips[, .(trip_id, trip_headsign, shape_id)])

# keeping trips CPTM L07-0 and CPTM L07-1
smaller_gtfs <- filter_by_trip_id(gtfs, c("CPTM L07-0", "CPTM L07-1"))
utils::object.size(smaller_gtfs)

head(smaller_gtfs$trips[, .(trip_id, trip_headsign, shape_id)])

unique(smaller_gtfs$shapes$shape_id)
```

We can see from the code snippet above that the function not only filters the `trips` table, but all other tables that contain a key that can be identified via its relation to `trip_id`. For example, since the trips `CPTM L07-0` and `CPTM L07-1` are described by the shapes `17846` and `17847`, respectively, these are the only shapes kept in `smaller_gtfs`.

The function also supports the opposite behaviour: instead of keeping the entries related to the specified identifiers, you can *drop* them. To do that, set the `keep` argument to `FALSE`:

```{r}
# dropping trips CPTM L07-0 and CPTM L07-1
smaller_gtfs <- filter_by_trip_id(
    gtfs,
    c("CPTM L07-0", "CPTM L07-1"),
    keep = FALSE
)
utils::object.size(smaller_gtfs)

head(smaller_gtfs$trips[, .(trip_id, trip_headsign, shape_id)])

head(unique(smaller_gtfs$shapes$shape_id))
```

And the specified trips (and their respective shapes as well) are nowhere to be seen. Now note that, since we are keeping many more entries in the second case, the resulting GTFS object, though smaller than the original, is much larger than in the first case.

The same logic applies to other filtering functions that you can use to filter a GTFS feed by `trip_id`, `route_id`, `shape_id`, `stop_id`, `route_type` or `service_id`.

## Filtering GTFS by day of the week or period of the day:

Some times, though, you might want to analyze only service levels on a certain days of the week or different periods of the day.

```{r}
# keeps entries related to services than run on Sunday AND Saturday
smaller_gtfs <- filter_by_weekday(gtfs, 
                                  weekday = c("sunday", "saturday"), 
                                  combine="and")

smaller_gtfs$calendar[, c("service_id", "sunday", "saturday")]
object.size(smaller_gtfs)

# keeps entries related to services than run EITHER Sunday or Saturday
smaller_gtfs <- filter_by_weekday(gtfs, 
                                  weekday = c("sunday", "saturday"), 
                                  combine="or")
smaller_gtfs$calendar[, c("service_id", "sunday", "saturday")]
object.size(smaller_gtfs)

```

Meanwhile, the `filter_by_time_of_day()` allows you to filter for example all services within a specified time period. 


```{r}
smaller_gtfs <- filter_by_time_of_day(gtfs, from = "14:00:00", to = "15:00:00")

# filter_by_time_of_day filters the frequencies table but doesn't filter the
# stop_times table because they're just templates
head(smaller_gtfs$frequencies)
head(smaller_gtfs$stop_times)

```


## Filtering GTFS by a spatial extent:

The `filter_by_sf()` function allows you to filter GTFS data inside or outside a given spatial extent. This functions takes a spatial `sf`/`sfc` object (or its bounding box) and keeps/drops the entries related to shapes and trips selected via a specified spatial operation. It may sound a bit complicated, but it's fairly easy to understand when shown. Let's create an auxiliary function to save us some typing:

```{r}
plotter <- function(gtfs,
                    geom,
                    spatial_operation = sf::st_intersects,
                    keep = TRUE,
                    do_filter = TRUE) {

  if (do_filter) {
    gtfs <- filter_by_sf(gtfs, geom, spatial_operation, keep)
  }

  shapes <- convert_shapes_to_sf(gtfs)
  trips <- get_trip_geometry(gtfs, file = "stop_times")
  geom <- sf::st_as_sfc(geom)

  ggplot() +
    geom_sf(data = trips) +
    geom_sf(data = shapes) +
    geom_sf(data = geom, fill = NA)

}
```

This function:

- Conditionally filters a GTFS object given a spatial object (called `geom`);
- Generates shapes' and trips' geometries as described in their respective tables;
- Generates a polygon from the bounding box;
- Plots all the `sf` objects cited above to show the effect of each `filter_by_sf()` argument in the final result.

Also, please note that our `plotter()` function takes the same arguments of `filter_by_sf()` (with the exception of `do_filter`, which is used to show the unfiltered data), as well as the same defaults.

Let's say that we want to filter GTFS data using the bounding box of the shape `68962`. Here's how the unfiltered data looks like, with the bounding box placed on top of it.

```{r}
bbox <- sf::st_bbox(convert_shapes_to_sf(gtfs, shape_id = "68962"))

plotter(gtfs, bbox, do_filter = FALSE)
```

By default `filter_by_sf()` (and `plotter()`, consequently) keeps all the data related to the trips and shapes that intersect the given geometry. Here's how it looks like:

```{r}
plotter(gtfs, bbox)
```

Alternatively you can also *drop* such data:

```{r}
plotter(gtfs, bbox, keep = FALSE)
```

You can also control which spatial operation you want to use to filter the data. This is how you'd keep the data that is contained inside the given geometry:

```{r}
plotter(gtfs, bbox, spatial_operation = sf::st_contains)
```

And, simultaneously using `spatial_operation` and `keep`, this is how you'd drop the data contained inside the geometry:

```{r}
plotter(gtfs, bbox, spatial_operation = sf::st_contains, keep = FALSE)
```

All filtering functions return a GTFS object readily available to be manipulated and analyzed using the rest of **gtfstools**' toolkit. For more information on how to use other functions made available by the package, please see the [introductory vignette](gtfstools.html).
