---
title: "Filtering GTFS feeds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Filtering GTFS feeds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

GTFS feeds are oftenly used to describe very large and complex public transport networks. These large files can become quite cumbersome to use, manipulate and move around, and it's not unfrequent that one wants to analyse a specific subset of the data.

**gtfstools** includes a few functions to filter GTFS feeds, thus allowing for faster and more convenient data processing. This vignette will introduce you to these functions and will cover their usage in detail.

We will start by loading the packages required by this demonstration into the current **R** session:

```{r, message = FALSE}
library(gtfstools)
library(ggplot2)
```

**gtfstools** includes the following filtering functions:

- `filter_by_trip_id()`
- `filter_by_route_id()`
- `filter_by_shape_id()`
- `filter_by_stop_id()`
- `filter_by_route_type()`
- `filter_by_sf()`

The first five work in a very similar fashion. You specify a vector of identifiers, and the function keeps (or drops, as we'll see soon) all the entries that are in any way related to this id. Let's see how that works using `filter_by_trip_id()`:

```{r}
path <- system.file("extdata/spo_gtfs.zip", package = "gtfstools")
gtfs <- read_gtfs(path)
utils::object.size(gtfs)

head(gtfs$trips[, .(trip_id, trip_headsign, shape_id)])

# keeping trips CPTM L07-0 and CPTM L07-1
smaller_gtfs <- filter_by_trip_id(gtfs, c("CPTM L07-0", "CPTM L07-1"))
utils::object.size(smaller_gtfs)

head(smaller_gtfs$trips[, .(trip_id, trip_headsign, shape_id)])

unique(smaller_gtfs$shapes$shape_id)
```

We can see from the code snippet above that the function not only filters the `trips` table, but all other tables that contain a key that can be identified via its relation to `trip_id`. For example, since the trips `CPTM L07-0` and `CPTM L07-1` are described by the shapes `17846` and `17847`, respectively, these are the only shapes kept in `smaller_gtfs`.

The function also supports the opposite behaviour: instead of keeping the entries related to the specified identifiers, you can *drop* them. To do that, set the `keep` argument to `FALSE`:

```{r}
# dropping trips CPTM L07-0 and CPTM L07-1
smaller_gtfs <- filter_by_trip_id(
    gtfs,
    c("CPTM L07-0", "CPTM L07-1"),
    keep = FALSE
)
utils::object.size(smaller_gtfs)

head(smaller_gtfs$trips[, .(trip_id, trip_headsign, shape_id)])

head(unique(smaller_gtfs$shapes$shape_id))
```

And the specified trips (and their respective shapes as well) are nowhere to be seen. Keep an eye that, since we are keeping many more entries in the second case, the resulting GTFS object, though smaller than the original, is much larger than in the first case.

The rest of the filtering functions all work very similarly, with the exception of one. `filter_by_sf()` takes a spatial `sf`/`sfc` object (or its bounding box) and keeps/drops the entries related to shapes and trips selected via a specified spatial operation. It may sound a bit complicated, but it's fairly easy to understand when shown. Let's create an auxiliary function to save us some typing:

```{r}
plotter <- function(gtfs,
                    geom,
                    spatial_operation = sf::st_intersects,
                    keep = TRUE,
                    do_filter = TRUE) {

  if (do_filter) {
    gtfs <- filter_by_sf(gtfs, geom, spatial_operation, keep)
  }

  shapes <- convert_shapes_to_sf(gtfs)
  trips <- get_trip_geometry(gtfs, file = "stop_times")
  geom <- gtfstools:::polygon_from_bbox(geom)

  ggplot() +
    geom_sf(data = trips) +
    geom_sf(data = shapes) +
    geom_sf(data = geom, fill = NA)

}

bbox <- sf::st_bbox(convert_shapes_to_sf(gtfs, shape_id = "68962"))

# unfiltered
plotter(gtfs, bbox, do_filter = FALSE)
```

